Stagnating State Support for Higher Education and Its Effect on Faculty Composition

### 2022-08-12 text/paper.tex
First draft.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
    Cornell University, Department of Economics.}
    \thanks{I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, and Francine Blau for various comments on the project's direction,
    as well as seminar participants at Cornell University (Autumn 202) for helpful
    discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/finances-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.}}
% Title
\title{Stagnating Public University Finances and Faculty Composition}
\date{\today}
\rhead{\today}
\lhead{Stagnating Public University Finances and Faculty Composition}
% Begin
\begin{document}
\clearpage \maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
\noindent
State support for public universities has stagnated the last thirty years, while enrollment has secularly increased.
At the same time, public universities have reduced the number of professors per student on tenure-track or with tenure, while increasing their employment of contingent lecturers.
This analysis relates changes in the faculty composition at public universities to the changes in per student university revenues from state appropriations, addressing endogeneity in state appropriation decisions by a shift-share instrumental variables approach.
An increase of university revenues of 10\%, via a change in state appropriations, increases the number of assistant professors per student by 6.6\% and full professors by 5.3\%, yet decreases the number of lecturers per student by 14.3\%.
Analysis of all the professors hired at Illinois public universities 2011-2021 implies that incumbent professors are not affected by the changes in university revenues, so that changes in faculty composition arises by changes in hiring practices at public universities.
These results show the compositional effects of stagnating state support for higher education, and raise questions about the direction that public education heads as these financial headwinds continue.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
\noindent
Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) the last three decades.
This decline has been shown to lead to worse education and later-life outcomes for students \citep{NBERw23736,NBERw27885}, yet it is not clear what mechanisms drive these effects or how public universities are affected as institutions.
I show that four-year, degree-granting public universities have experienced falling state funding, leading to falling revenues per student, effecting the composition of professors within public universities in general.
Falls in public university revenues from the state, instrumented by state-level finance shocks, lead to a fall in the number of tenure-track or tenured professors per student within a university, an increase in the number of lecturers, and overall fall in the number of faculty per student.
US private universities were not exposed to such financial constraints during this same time period, and do not exhibit the substitution away from tenure-track and tenured professors, so that the stagnating state support for public universities has implications for the structure of higher education instruction and research in the US.

US Universities are widely considered the highest performing in the world, yet there are consequential differences between its universities that operate in the private sector and those established by state governments.
public universities are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average PU received around \$11,600 per enrolled student in 1990, yet only \$8,300 per enrolled student in 2021.\footnote{
    2021 figures are relatively similar to those for year preceding, despite the Covid-19 pnademic and associated financial shocks, since 2021 refers to the academic year 2019-2021 and so the yearly budget was decided in 2019.
    This paper does not directly analyse budget shocks related to the Covid-19 pandemic, or any other events in 2021 and following.
}
This fall is driven by stagnating funding provided by state governments, while total enrollment among public universities rose by 46\% over the same time period.
At the same time, the number of professors per student at public universities stayed relatively stable at 0.045, yet fell from around 0.04 to around 0.35 professors per student at public universities, driven by a fall of over 20\% in the ratio of associate and full (i.e. tenured) professors per student at public universities.\footnote{
    \autoref{fig:fte-perprof} and \autoref{sec:trends} document these trends between private and public universities in full detail.
}

I use a variant of the \cite{NBERw23736,NBERw27885} shift-share instrument for PU revenues from shocks to state appropriations to identify the effect on the faculty composition.
State appropriations for public universities are plausibly endogenous to many outcomes: state governments decide on yearly budgets for their higher education sector, and this process can be influenced by local financial conditions, or even public (and political representatives') perceptions of the state's higher education system.
The shift-share instrument interacts reliance on state funding in a base period with the yearly total appropriations per student in the entire state, to exploit both changes in state support for higher education and how much each university relies on that support.

Findings show that negative changes in university revenues affects faculty composition at public universities, away from tenured and tenure-track professors towards non-tenure track lecturer positions.
A reduction of 10\% in university revenues, via a shock to state appropriations, leads to a fall in 6.6\% in the number of assistant professors per student within a university, 5.3\% for full (tenured) professors, and a reduction in 14.3\% for lecturers.
Additionally, I use individual level data on professors who were hired at Illinois public universities 2011-2021 to investigate whether the shocks to university revenues affected incumbent professors at public universities.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system, which implies that changes in faculty composition came about by changes in the composition of hiring new professors at the university level.

\cite{NBERw23736} first study the effect of (changes in) public universities' finances on student outcomes by isolating changes in public universities' expenditures on student instruction, addressing endogeneity of expenditures decisions by instrumenting expenditures by an instrument for state appropriations and tuition caps.
\cite{NBERw23736} find that increases in PU spending (via state appropriations) increases enrolment and degree completion among students, but corresponding changes in attendance price (via tuition caps/freezes) do not have an effect.
\cite{chakrabarti2018effect,NBERw27885} use the same instrument for PU revenues at the state level, combined with richer information on students' individual-level outcomes, to show that increases in state appropriations lead to lower time to completion and debt levels among students later in life.
\cite{bound2019public} use another variant of the same approach to show that state-level higher education spending cuts induced public universities to shift toward tuition as their primary source of revenue, and shift institutional resources towards students who pay higher tuition rates.
In a similar vein, \cite{bound2007cohort} document variation in per student state funding for higher education resulting from yearly changes in a state's student-age population; they find that lower per student funding lead to lower higher education completion rate among students.

University professors the core component of this country's educational and research/innovation university system, yet their composition within universities and its relationship with higher education finances has so far not been studied.
\cite{brown2014endowment} presents the closest example by studying how university endowments react to negative financial shocks, exhibiting an association between negative endowment shocks and a fall in tenure-system faculty employed at the university.
\cite{abe2015implications} present a model for university administrator decision-making, and posit that a large component of variation in faculty salaries arises from university politics, and not only economic factors such as outside.
\cite{johnson2009jep,NBERc13879} both empirically document allocation of faculty between research and instruction, and \cite{hemelt2021math} document differences in instructional cost (per student) between departments, primarily thanks to department differences in class size and faculty salaries.
\cite{turner2014impact} document how universities reacted to the Great Recession of 2008, a severe financial shock to state finances, including implementing hiring freezes.
Indeed, Cornell University\footnote{
    Cornell University is, indeed, only partially public, so that this information is not free information, and cannot be presented with aggregate or verified sources of public data on such salaries.
} implemented both hiring freezes and even a nominal salary reduction for professors, in anticipation of oncoming financial shock in early 2021.\footnote{
    The salary cut was not permanent, as the oncoming financial shock turned out to not be as serious as projected, and the salary cut was returned to professors in the year 2021.
}

The number of professors, in absolute terms as well as relative to number of students in the university, may be particularly important for quality of instruction and thus educational outcomes.
\cite{angrist1999using} first causally show that reducing class size induces an increase in test scores for school-age children, yet the magnitude of the effect is not as clear in the university setting.
\cite{bandiera2010heterogeneous} use a fixed effect model, combined with multiple individual observations in a long panel, to find that UK university students perform worse academically in particularly large classes, and the effect is particularly largest for students at the top of the academic distribution.
And for the composition of faculty teaching university classes, 
\cite{bettinger2010does,figlio2015tenure} find US students have better enrollment and learning outcomes from courses taught by contingent or adjunct (i.e. not tenured or on tenure-track) professors (relative to tenured or tenure-track).
While \cite{ehrenberg2005tenured} observe a negative association between utilisation of non-tenured instructors and student graduation rates.
My approach focuses on count of professors per student at universities, so does not directly observe individual class-sizes, yet presents this as a possible mechanism through which the negative effects on student outcomes that falls in PU finance \cite{NBERw23736,NBERw27885} document.

This analysis finds that faculty composition is meaningfully affected by recent stagnation in state support for higher education.
Analysis of outcomes for incumbant professors in Illinois implies that the explaining mechanism is changes to hiring by public universities in response to the increased financial constraints.
Previous findings from the literature on the effectiveness of tenured and tenure-track faculty versus contingent faculty in the areas of research and teaching.
Combined with this analysis, these findings provide a plausible mechanism for the deteriorating student ouctomes shown to be a result of stagnating finances at public universities \citep{NBERw23736,NBERw27885}.

@@

\subsection{Data Description}
The data used in this analysis come from two primary sources: \citet[IPEDS]{ipeds} regarding institutional information on finances and enrollment, and \citet[IBHED]{ibhed} for a panel of individual-level of every Professor employed by the Illinois PU system.

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    This statement means that IPEDS does not necessarily cover the universe of US higher education institutions, yet thin practice every PU and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e. 1987 refers to the academic year that ran August 1986 to July 1987.
}
and provide information on the total revenues a university received from every source (including state governments), enrollment number, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990-2017, and manually combine with raw NCES data regarding the year 2018-2021 for all relevant variables.
    Figures for enrollment, faculty counts and salaries come from the raw NCES version of IPEDS for all years to address inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to four-year institutions, as these institutions adhere to a standardised concept of faculty profile -- that is, tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised in four-year, degree-granting institutions.
Additionally, for-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enroll at a four-year institution \citep{mountjoy2022}.\footnote{
    Furthermore, \cite{mountjoy2022} documents the effects of expanding two-year higher education access in the US separately for students who would have and those who would not have otherwise have attended college.
    The high rate of eventual enrollment at four-year institutions among two-year students explains the negative impact of increased access to two-year institutions on students who are diverted away from four-year institutions.
}

Importantly, IPEDS reports the count of professors employed by category, as well as total salary expenditures by category.
This gives a resulting panel data-set, where each row represents a university-year, and includes columns for university finances, plus total count and average salary\footnote{
    Real salary is computed by scaling nominal salary to 2021 dollars by the CPI-U.
} for professors by category (lecturer, assistant, tenured, total).
\autoref{tab:ipeds-summary} presents summary statistics for relevant variables in the panel of PU-years.

@@

IPEDS provides information at the university level, yet lacks information on the distribution of salaries and professor count within a university.\footnote{
    IPEDS provides total paid to salary and total faculty employed per university-year, so that I can investigate average professor salary per university-year with IPEDS data.
    Unfortunately, it seems that this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period \citep{aau2021survey}.
}
To investigate the distribution, I integregate individual-level data for every PU professor in the state of Illinois between the years 2010-2021.
IBHED host the information;
Public Act 96-0266 (effective 1 January 2010) requires that each university report base salary and benefits all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the eight Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
This publicly available data provides the basis to build a panel of Illinois PU professors 2010-2021; I define a professor as an individual observation by their first plus last name and university pairing, and link this database to IPEDS information on finances for their employing institution.

The analysis sample represents 16,932 professors in the year 2010 and 15,352 in the year 2021, with summary statistics presented in \autoref{tab:illinois-summary}.
Additionally, further analysis using the first year of a professor's employment focuses on the subset of professors with observed year of hiring between the years 2011 and 2021, representing 1,778 professors in 2011, and 9,099 in 2021.\footnote{
    The reasons for restricting to the sample with identified year of hire in the range 2011-2021 is explained in \autoref{sec:iv-model-indiv}.
}
Summary statistics for this group, which over-represents lecturers compared to assistant and full professors, can be seen in \autoref{tab:illinois-summary-rolling}.

@@

\subsection{Trends in Finances, Enrollment, and Faculty Composition}
\label{sec:trends}
A state government in the US plans an annual budget a couple of years ahead of the fiscal year, with the legislature approving a budget request put forth by the governor's office.\footnote{
    \cite{NBERw23736} present a full discussion of the state appropriation decision-making process, drawing on administrative records originally analysed by \cite{parmley2009state}.}
This process leads to yearly variation in state appropriations not seen in other revenues sources (such as federal government appropriations), since US states differ in their rate of (financial) support for higher education, and are subject to fiscal constraints.
US states governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (caused by changing economic conditions or otherwise) necessarily affect state expenditures.
State appropriations to higher education are a particularly attractive area of state spending to absorb such shocks to state finances \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describes the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends.
}
Additionally, there is yearly state variation in the number of higher education students \citep{turner2014impact}, so that per-student state appropriations vary on multiple dimensions.

\autoref{fig:funding} exhibits the trends for the mean of US public universities between the years 1990-2021.
We see a rise in total revenues received by public universities (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university. 
Non-institutional revenues refer to revenues coming from external sources, not raised by internal university activities;\footnote{
    US universities raise revenues from many internal activities, including sales of nondegree related educational courses to publishing deals.
    The non-institutional revenue measure is constructed to consider only revenues resulting from external decisions, including state, local, and federal appropriations and tuition revenues.
    Rates of tuition are considered external since most state governments play a large role in deciding tuition rices for public universities \citep{NBERw23736}.
}
non-institutional revenues rose over this time period almost exclusively thanks to the rise in tuition revenues.
At the same time, total state appropriations stagnated at around \$100 million per year-university for 1990-2008, fell slightly around 2008 and have not recovered.
While public universities experienced a financial stagnation, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021 with no corresponding decline in any specific component.

@@

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Total Student Enrollment, by University Sector, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total.}
        \includegraphics[width=\textwidth]{figures/enrollment-total.png}
        \label{fig:enrollment-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Mean, University Level.}
        \includegraphics[width=\textwidth]{figures/enrollment-mean.png}
        \label{fig:enrollment-mean}
    \end{subfigure}
    \label{fig:enrollment}
\end{figure}
At the same time, student enrollment at public universities rose precipitously.
6.2 million students were enrolled in public universities in 1990, and this number rose by 47\% to 9.1 million, with most of the increase occurring over the years 2000-2021.
\autoref{fig:enrollment} shows that total enrollment at public universities has also risen over the same time period, but not as drastic in either relative or absolute terms; the mean public university also grew over this time period, from 9,800 students in 1990 to 11,800 in 2021.
This means that per student mean revenues per student (seen in \autoref{fig:mean-funding-fte}) have stagnated across all measures, and particularly fallen from \$11,000 per student in 1990 to less than \$8,000 per student in 2021.
\begin{figure}[!h]
    \centering
    \singlespacing
    \caption{Total Professor Count per Student, by University Sector, Professor Appointment, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Lecturer}
        \includegraphics[width=\textwidth]{figures/lecturer-fte-perprof.png}
        \label{fig:lecturer-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Assistant.}
        \includegraphics[width=\textwidth]{figures/assistant-fte-perprof.png}
        \label{fig:assistant-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Full.}
        \includegraphics[width=\textwidth]{figures/full-fte-perprof.png}
        \label{fig:full-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{All.}
        \includegraphics[width=\textwidth]{figures/all-fte-perprof.png}
        \label{fig:all-fte-perprof}
    \end{subfigure}
    \label{fig:fte-perprof}
\end{figure}

\autoref{fig:fte-perprof} documents the divergence in faculty composition (per student) between the average private and public university 1990-2021.
Public universities start with  a higher baseline of around 4.5 professors per 100 students, and exhibits yearly variation of less than 0.5 professor per 100 students over the thirty years.
Public universities start with 3.9 professors per student, and this number falls to 3.5 primarily in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for public universities: in 2021 the mean PU has 6 fewer full professors per student than the mean private university.
At the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990 yet have been utilised by public universities at a higher rate since.

@@

The state of Illinois particularly exemplifies the stagnation in state support for higher education that the rest of the nation has experienced 1990-2021.
\autoref{fig:illinois-funding} shows the trends for revenues among the seven Illinois public university campuses, including the falling share of state appropriations and substitution towards tuition revenue.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
    Cornell University, Department of Economics.}
    \thanks{I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, and Francine Blau for various comments on the project's direction,
    as well as seminar participants at Cornell University (Autumn 202) for helpful
    discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/finances-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.}}
% Title
\title{Stagnating State Support for Higher Education and Its Effect on Faculty Composition}
\date{\today}
\rhead{\today}
\lhead{State Support for Higher Education and Faculty Composition}
% Begin
\begin{document}
\clearpage \maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
\noindent
State support for higher education has stagnated on a per student level the last thirty years, while public universities have reduced the number of tenure-track or with tenured professors per student, and increased their employment of contingent lecturers.
This analysis relates these changes in the faculty composition at public universities to the fall in per student university revenues from state appropriations.
A shift-share instrumental variables approaches addresses endogeneity in decisions for state appropriations to higher education, by exploiting the yearly change in state support with an institution's reliance on that support.
A decrease in state funding of 10\%, via a shock to state appropriations, decreases the number of assistant professors per student at a public university by 1.4\% and full professors by 1.2\%, yet increases the number of lecturers per student by 4.4\%.
Analysis of all the professors hired at Illinois public universities 2011-2021 shows that incumbent professors are not affected by the changes in state support, implying that these changes in faculty composition arose by lower rate of hiring of tenure-track and tenured professors at public universities.
These results show one of the long-term effects of stagnating state support for higher education, and raises questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
\noindent
Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) the last three decades.
This decline has been shown to lead to worse education and later-life outcomes for students \citep{NBERw23736,NBERw27885}, yet it is not clear what mechanisms drive these effects, or how public universities are affected as institutions.
I show that four-year, degree-granting public universities have experienced falling state funding per student, and this affected their composition of faculty.
Falls in public university revenues from the state, instrumented by state-level finance shocks, lead to a fall in the number of tenure-track or tenured professors per student within a university, an increase in the number of lecturers, and overall fall in the number of faculty per student.
US private universities were not exposed to such financial constraints during this same time period, and do not exhibit the substitution away from tenure-track and tenured professors, so that the stagnating state support for public universities has implications for the wider structure of higher education instruction and research in the US.

US universities are widely considered the highest performing in the world, yet there are consequential differences between its universities that operate in the private sector and those established by state governments.
Public universities are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average public university received around \$11,600 per enrolled student in 1990, yet only \$8,300 per enrolled student in 2021.
This fall is driven by stagnating funding provided by state governments, while enrolment among public universities rose by 46\% over the same time period.
At the same time, the number of professors per student at public universities stayed relatively stable at 0.045, yet fell from around 0.04 to around 0.35 professors per student at public universities, driven by a fall of over 20\% in the ratio of associate and full professors per student at public universities.

I use a variant of the \cite{NBERw23736,NBERw27885} shift-share instrument for state funding from shocks to appropriations for the entire state to identify the effect on the faculty composition.
State appropriations for public universities are plausibly endogenous to many outcomes: state governments decide on yearly budgets for their higher education sector, and this process can be influenced by local financial conditions, or even public perceptions of the state's higher education system.
The shift-share instrument interacts reliance on state funding in a base period with the yearly total appropriations per student in the entire state, to exploit both changes in support for higher education and how much each university relies on that support.

Findings show that negative changes in state funding affects faculty composition at public universities, away from tenured and tenure-track professors towards non-tenure track lecturer positions.
A reduction of 10\% in state funding, via a shock to state appropriations, leads to a fall in 1.4\% in the number of assistant professors per student within a university, 1.2\% for full (tenured) professors, and an increas in 4.4\% the number of lecturers per student.
Additionally, I use individual level data on professors at Illinois public universities over 2010-2021 to investigate whether the shocks to state funding affected incumbent professors at public universities.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system, yet lecturers' first year slary decreased by 3.8\% in response to a fall in state funding of 10\%.
These findings imply that changes in faculty composition came about by changes in the composition of hiring new professors at the university level.

\cite{NBERw23736} first study the effect of (changes in) public universities' finances on student outcomes by isolating changes in public universities' expenditures on student instruction, addressing endogeneity of expenditures decisions by instrumenting expenditures with state appropriation shocks and limits on tuition price.
\cite{NBERw23736} find that increases in public university spending (via appropriation shocks) increases enrolment and degree completion among students, but corresponding changes in attendance price (via limits on tuition price) do not have an effect.
\cite{chakrabarti2018effect,NBERw27885} use the same instrument for public university revenues at the state level, combined with rich information on students' individual-level outcomes, to show that increases in state appropriations lead to degree lower completion time and later-life student debt.
\cite{bound2019public} use another variant of the same approach to show that state-level higher education spending cuts induced public universities to shift toward tuition as their primary source of revenue, and shift institutional resources towards students who pay higher tuition rates.
In a similar vein, \cite{bound2007cohort} document variation in per student state funding for higher education resulting from yearly changes in a state's student-age population; they find that lower per student funding lead to lower higher education completion rate among students.

Faculty are a core component of this country's educational and research-innovation university system, yet their composition within universities, and its relationship with higher education finances has so far not been studied.
\cite{brown2014endowment} presents the closest example by studying how university endowments react to negative financial shocks, exhibiting an association between negative endowment shocks and a fall in tenure-system faculty employed at private universities.
\cite{abe2015implications} present a model for university administrator decision-making, and posit that a large component of variation in faculty salaries arises from university politics, and not only economic factors such as outside options.
\cite{johnson2009jep,NBERc13879} both empirically document allocation of faculty between research and instruction, and \cite{hemelt2021math} document differences in instructional cost (per student) between departments, primarily thanks to department differences in class size and faculty salaries.
\cite{turner2014impact} document how universities reacted to the Great Recession of 2008, a severe financial shock to state finances and support for higher education.

The number of professors, both in absolute terms and relative to number of students, may be particularly important for quality of instruction and thus educational outcomes.
\cite{angrist1999using} first causally show that reducing class size (via religious ruling) induces an increase in test scores for school-age children, yet the magnitude of this effect is not as clear in the higher education setting.
\cite{bandiera2010heterogeneous} use a fixed effect model, combined with multiple individual observations in a long panel, to find that UK university students perform worse academically in particularly large classes, and the difference is largest for students at the top of the academic distribution.
For the composition of faculty teaching university classes, \cite{bettinger2010does,figlio2015tenure} find US students have better enrolment and learning outcomes from courses taught by contingent or adjunct (i.e. not tenured or on tenure-track) professors (relative to tenured or tenure-track).
\cite{ehrenberg2005tenured} observe a negative association between utilisation of non-tenured instructors and student graduation rates.
My approach focuses on count of professors per student at universities, so does not directly observe individual class-sizes, yet presents considers this a possible mechanism through which stagnating state support for higher education negatively affects student outcomes \citep{NBERw23736,NBERw27885}.

@@

\subsection{Data Description}
The data used in this analysis come from two primary sources: \citet[IPEDS]{ipeds} regarding institutional information on finances and enrolment, and \citet[IBHED]{ibhed} for data on every professor in the Illinois public university system.

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    This statement means that IPEDS does not necessarily cover the universe of US higher education institutions, yet in practice every public university and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e. 1987 refers to the academic year that ran August 1986 to July 1987.
}
and provide information on the total revenues a university received from every source (including state governments), enrolment number, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990-2017, and manually combine with raw NCES data regarding the year 2018-2021 for all relevant variables.
    Figures for enrolment, faculty counts and salaries come from the raw NCES version of IPEDS for all years to address inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to public, four-year, degree-granting institutions, as these institutions adhere to a standardised concept of faculty profile, where tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised.
For-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enrol at a four-year institution \citep{mountjoy2022},\footnote{
    Furthermore, \cite{mountjoy2022} documents the effects of expanding two-year higher education access in the US separately for students who would have and those who would not have otherwise have attended college.
    The high rate of eventual enrolment at four-year institutions among two-year students explains the negative impact of increased access to two-year institutions on students who are diverted away from four-year institutions.
}
so that these institutions are not considered.

IPEDS reports the count of professors employed by position, as well as total salary expenditures by position.
This gives a resulting panel data-set, where each row represents a university-year, and includes columns for university finances, plus total count and average salary\footnote{
    Real salary is computed by scaling nominal salary to 2021 dollars by the CPI-U.
} for professors by position (lecturer, assistant, tenured, total).
\autoref{tab:ipeds-summary} presents summary statistics for relevant variables in these data.

@@

IPEDS provides information at the university level, yet lacks information on the distribution of salaries and professor count within a university.\footnote{
    IPEDS provides total paid to salary and total faculty employed per university-year, so that I can investigate average professor salary per university-year with IPEDS data.
    Unfortunately, this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period \citep{aau2021survey}.
    Similarly, IPEDS reports of the average salary paid to professors at private institutions also disagree with other sources, so are not considered in this analysis.
}
To investigate the distribution, I integrate individual-level data for every public university professor in the state of Illinois between the years 2010-2021.
IBHED hosts the information;
Public Act 96-0266 (effective 1 January 2010) requires that each university report base salary and benefits all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the nine Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
This publicly available data provides the basis to build a panel of Illinois public university professors 2010-2021; I define a professor as an individual by their first plus last name and university pairing, and link this database to IPEDS regarding finances for their employing institution.

The analysis sample represents 16,932 professors in the year 2010 and 15,352 in the year 2021, with summary statistics presented in \autoref{tab:illinois-summary}.
Additionally, further analysis using the first year of a professor's employment focuses on the subset of professors with observed year of hiring between the years 2011 and 2021, representing 1,778 professors in 2011, and 9,099 in 2021.\footnote{
    Summary statistics for this group, which over-represents lecturers compared to assistant and full professors, can be seen in \autoref{tab:illinois-summary-rolling}.
    The reasons for restricting to the sample with identified year of hire in the range 2011-2021 is explained in \autoref{sec:iv-model-indiv}.
}

@@

\subsection{Trends in Finances, Enrolment, and Faculty Composition}
\label{sec:trends}
A state government plans an annual budget a couple of years ahead of the fiscal year, and the legislature chooses to approve or reject a budget request put forth by the governor's office.\footnote{
    \cite{NBERw23736} present a full discussion of the state appropriation decision-making process, drawing on administrative records originally analysed by \cite{parmley2009state}.}
US state governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (caused by changing economic conditions or otherwise) necessarily affect state expenditures.
This process leads to yearly variation in state appropriations not seen in other revenues sources (such as federal government appropriations), since US states differ in their support for higher education, and are subject to fiscal constraints.
State appropriations to higher education are a particularly attractive area of state spending to absorb such shocks to state finances \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describes the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends.
}
Additionally, there is yearly variation in the number of higher education students in each state \citep{turner2014impact}, so that per-student state appropriations vary on multiple dimensions.

\autoref{fig:funding} exhibits the trends for the mean public university for the years 1990-2021.
We see a rise in total revenues received by public universities (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university.
At the same time, total state appropriations stagnated at around \$100 million per year-university for 1990-2008, falling around 2008 and have not recovered ever since.
While public universities experienced a stagnation in state support, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021, experiencing no corresponding decline in any specific component.

@@

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Total Student enrolment, by University Sector, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total.}
        \includegraphics[width=\textwidth]{figures/enrollment-total.png}
        \label{fig:enrollment-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Mean, University Level.}
        \includegraphics[width=\textwidth]{figures/enrollment-mean.png}
        \label{fig:enrollment-mean}
    \end{subfigure}
    \label{fig:enrolment}
\end{figure}

At the same time, student enrolment at public universities rose precipitously.
6.2 million students were enrolled in public universities in 1990, and this number rose by 47\% to 9.1 million, with most of the increase occurring over the years 2000-2021.
\autoref{fig:enrolment} shows that total enrolment at private universities has also risen over the same time period, but not as drastic in either relative or absolute terms; the mean private university grew from 9,800 students in 1990 to 11,800 in 2021.
This means that revenues per student have stagnated across all measures (seen in \autoref{fig:mean-funding-fte}), and particularly fallen from a mean of \$11,000 per student in 1990 to less than \$8,000 per student in 2021.

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Total Professor Count per Student, by University Sector, Professor Appointment, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Lecturer}
        \includegraphics[width=\textwidth]{figures/lecturer-fte-perprof.png}
        \label{fig:lecturer-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Assistant.}
        \includegraphics[width=\textwidth]{figures/assistant-fte-perprof.png}
        \label{fig:assistant-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Full.}
        \includegraphics[width=\textwidth]{figures/full-fte-perprof.png}
        \label{fig:full-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{All.}
        \includegraphics[width=\textwidth]{figures/all-fte-perprof.png}
        \label{fig:all-fte-perprof}
    \end{subfigure}
    \label{fig:fte-perprof}
\end{figure}

\autoref{fig:fte-perprof} documents the divergence in faculty composition (per student) between the average private and public university 1990-2021.
Private universities start with  a higher baseline of around 4.5 professors per 100 students, and exhibited yearly variation of less than 0.5 professor per hundred students over the thirty years.
Public universities start with 3.9 professors per hundred students, and this number falls to around 3.5 primarily in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for public universities: in 2021 the mean public university has 6 fewer full professors per hundred students than the mean private university.
Over the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990 yet have been utilised by public universities at a higher rate since.

@@

The state of Illinois experienced the same stagnation in state support for higher education that the rest of the nation has experienced 1990-2021.
\autoref{fig:illinois-funding} shows the trends for revenues among the seven Illinois public university campuses, including the falling share of state appropriations and substitution towards tuition revenue.
